---
title: "Getting Started with EpigenicR"
author: "Nimra Ali"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with EpigenicR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE  # Set to TRUE when toy data is available
)
```

## Introduction

EpigenicR provides tools for processing, analyzing, and visualizing epigenomic data from ChIP-seq, methylation, and chromatin state experiments. This vignette demonstrates the basic workflow for using EpigenicR.

## Installation

See the [INSTALL.md](../INSTALL.md) file for detailed installation instructions.

```{r install, eval=FALSE}
# Install from local source
devtools::install_local("/path/to/EpigenicR")
```

## Load the package

```{r load}
library(EpigenicR)
library(dplyr)
library(ggplot2)
```

## Workflow

### 1. Extract Metadata from BigWig Files

The first step is typically to extract metadata from your BigWig filenames. EpigenicR expects filenames in a specific format:

`projectID_batch_marker_rerunID_sampleID_replicate.genome.scaling.bw`

For example: `project1_batch1_H3K4me3_rerun_sample1_rep1.hg38.scaled.bw`

```{r metadata}
# Example BigWig filenames
bw_files <- c(
  "project1_batch1_H3K4me3_rerun_sample1_rep1.hg38.scaled.bw",
  "project1_batch1_H3K4me3_rerun_sample1_rep2.hg38.scaled.bw",
  "project1_batch1_H3K9me3_rerun_sample2_rep1.hg38.scaled.bw",
  "project1_batch1_H3K9me3_rerun_sample2_rep2.hg38.scaled.bw"
)

# Extract metadata
metadata <- create_metadata_df(bw_files = bw_files)
print(metadata)
```

### 2. Download Genomic Annotations

EpigenicR provides functions to download and process genomic annotations:

```{r annotations, eval=FALSE}
# Download GTF and create BED files for genes and TSS regions
annotation_paths <- ensure_gtf_and_beds(
  gtf_file = "~/data/gencode.v38.annotation.gtf",
  gtf_url = "https://ftp.ebi.ac.uk/pub/databases/gencode/Gencode_human/release_38/gencode.v38.annotation.gtf.gz",
  genes_bed = "~/data/genes.hg38.bed",
  tss2k_bed = "~/data/genes_tss_2kb.hg38.bed"
)

# Download ChromHMM chromatin state annotations
download_chromhmm_annotations(
  annotations = c("E107_15_coreMarks_hg38lift_mnemonics.bed"),
  dest_dir = "~/data/chromHmm_annotations/"
)
```

### 3. Quality Control Visualization

Generate QC plots to assess data quality across markers and replicates:

```{r qc_data}
# Create example QC statistics
qc_data <- data.frame(
  marker = rep(c("H3K4me3", "H3K9me3", "H3K27me3"), each = 6),
  replicate = rep(paste0("rep", 1:3), 6),
  sample_id = rep(c("sample1", "sample2"), each = 9),
  condition = rep(c("Control", "Treatment"), each = 9),
  total_reads = runif(18, 1e6, 5e6),
  mapped_reads = runif(18, 8e5, 4.5e6),
  library_size = runif(18, 7e5, 4e6),
  frip_score = runif(18, 0.1, 0.5)
)
```

```{r qc_plot_static}
# Create static ggplot2 figure
qc_plot <- plot_qc_stats(
  data = qc_data,
  condition = "Control",
  stats = c("total_reads", "mapped_reads", "frip_score"),
  marker_levels = c("H3K4me3", "H3K9me3", "H3K27me3"),
  engine = "ggplot",
  legend_position = "bottom",
  ncol = 3
)
print(qc_plot)
```

```{r qc_plot_interactive}
# Create interactive plotly version
qc_interactive <- plot_qc_stats(
  data = qc_data,
  condition = "Control",
  stats = c("total_reads", "mapped_reads", "frip_score"),
  engine = "plotly",
  legend_position = "bottom",
  ncol = 3
)
qc_interactive
```

### 4. Working with EPK Objects

EPK (Epigenomic Package) objects bundle together MultiAssayExperiment data with QC statistics, enrichment results, and provenance information.

```{r epk_structure, eval=FALSE}
# EPK objects are typically created from processed BigWig data
# They contain:
# - mse: MultiAssayExperiment with multiple genomic feature sets
# - tables: QC statistics and summary tables
# - enrichment_results: Enrichment analysis results
# - provenance: Creation timestamp and session info

# Load an existing EPK object
epk <- readRDS("path/to/project.epk.rds")

# Print summary
print(epk)

# Access components
mse <- epk$mse
stats_summary <- epk$tables$stats_summary
enrichment <- epk$enrichment_results
```

### 5. Sample Correlation Analysis

Compute sample-sample correlations to assess replicate consistency:

```{r correlation, eval=FALSE}
# Compute correlations for all markers in the protein_coding experiment
cor_matrices <- compute_all_cor(
  mse = epk$mse,
  exp_name = "protein_coding",
  method = "pearson",
  transform = "log1p"
)

# Access correlation matrix for a specific marker
cor_h3k4me3 <- cor_matrices$H3K4me3

# Visualize correlation heatmap
library(ComplexHeatmap)
Heatmap(cor_h3k4me3, 
        name = "Correlation",
        col = circlize::colorRamp2(c(0, 1), c("white", "red")))
```

## MultiAssayExperiment Structure

Each EPK object contains a `MultiAssayExperiment` with experiments for different genomic feature types:

- **protein_coding**: Protein-coding genes
- **cpg**: CpG islands
- **tss_2k**: TSS ±2kb regions
- **vista_enhancer**: VISTA enhancers
- **E107_chromatin_state**: Chromatin state segments

Each experiment is a `SummarizedExperiment` with assays for each epigenomic marker (H3K4me3, H3K9me3, 5mC, etc.).

```{r mae_structure, eval=FALSE}
# Access experiments
experiments <- MultiAssayExperiment::experiments(epk$mse)

# Get protein coding genes data
se_genes <- experiments$protein_coding

# Access H3K4me3 data for protein coding genes
h3k4me3_matrix <- SummarizedExperiment::assay(se_genes, "H3K4me3")
```

## Extracting Marker Names

For flexible handling of sample IDs that contain marker information:

```{r extract_markers}
# Extract marker names from sample IDs
sample_ids <- c("Sample1_H3K4me3_Rep1", "Sample2_H3K9me3_Rep2", "Sample3_5mC_Rep1")
markers_list <- c("H3K4me3", "H3K9me3", "5mC", "H3K27me3")

extracted <- sapply(sample_ids, extract_marker_names, markers = markers_list)
print(extracted)
```

## Next Steps

- Explore the individual function documentation using `?function_name`
- See the [README.md](../README.md) for more examples
- Check out the full workflow in the `epk.R` file for creating complete EPK objects

## Session Info

```{r session_info}
sessionInfo()
```
